Unveiling Robustness Threats by Mining the Exception Stack Traces of over 600 Android Projects

### 2014-09-05 android-stacks.tex
threat

Along the recent years, we have witnessed an astonishing increase in the number of
mobile applications. If on the one hand such applications extend phones capabilities 
far beyond of the basic calls and textual messages, on the other hand
they have to face an increasing number of threats to application robustness
 arising from: faults on underlying middleware and hardware (e.g., GPS receiver, camera, sensors);
faults on third party services and libraries; compatibility issues ~\cite{McDon13}; 
 memory and battery restrictions~\cite{Zhang12}; among others.

@@

Studies have shown, however, that the exception-related code is generally poorly understood and the
 less tested parts of the system~\cite{miller1997issues,Robil00,shah2010understanding, 
garcia2007extracting,garcia2001comparative,cabral2007exception,coelho2011unveiling}.
As a consequence they may affect the system the other way around.
The exception-related code may favor the introduction of failures such as 
uncaught exceptions~\cite{jo2004uncaught, zhang2012amplifying} - 
which can lead to system crashes, making the system even less robust~\cite{coelho2011unveiling}.

@@

 \item  A multitude of programming mistakes were reported as the main causes of exception stack traces:
 27,71\% of the analyzed exception stack traces reported a java.lang.NullPointerException the root cause;
and  51,96\% of all projects reported at least one exception stack trace on which the NullPoiterException
was the root cause.

@@

Such outcomes point to threats not only to the developement of robust Android apps, 
but also to the developement of any robust Java-based system. For instance, in the absence of the 
``exception specification" of  a third-party code, it is difficult or 
even enfeasible for the developer to protect the code against ``unforeseen" exceptions. 
Since, in such cases the client usually do not have access to the source code, the undocumented exceptions may 
remain uncaught and lead to system crashes. Moreover, the cross-type exception wrappings 
(e.g., OutOfMemoryError is wrapped in a checked Exception) 
may lead to unintended handling problems - trying to handle an instance of OutOfMemoryError ``hiden" in 
a checked exception may lead the program to an unpredictable state.

@@

%The Android platform is mostly implemented in Java and uses Java Native Invocation (JNI) to access native C/C++ code. 
%And as mentioned before, it reuses the embedded Java Exception model to detect and handle 
%abnormal computation states.

The Android Application Framework supports an event-driven development model. 
Among the basic elements that comprise an Android application, we can cite the
 Activity class. The Activity is the main class of an Android app (there is usualy one
Activity per application window). Each Activity defines a set of hook methods  
(e.g., OnCreate(), onPause()) that are dispatched automatically by the Android 
framework (by inversion of control), in response to infrastructure and application events.

The Android Application Framework also provides several supporting APIs, such as 
the APIs to support concurrent development (e.g. AsyncTask). 
Such APIs particularly useful due to the high demand for responsiveness 
in mobile environment (mobile apps can easily become 
unresponsive due to resource limitation and excessive network access).

@@

In Java, an exception is represented according to the class hierarchy shown in
Figure~\ref{fig:exchier}.  According to it every exception is an
instance of the Throwable class, and can be of three kinds: the checked exceptions
(extends Exception), the runtime exceptions (extends RuntimeException) and errors
(extends Error)~\cite{gosling2000java}. The checked exception received such name
 because they must be declared on the method's exception interface (i.e., a list of exceptions that a method 
might raise during its execution) and the compiler statically checks if
 appropriate handlers are provided within the system.
Both runtime exceptions and errors are also known as unchecked exceptions, because 
they do not need to be specified on method exception interface and do not trigger any 
compile time checking.

By convention an Error represents an unrecoverable condition which usually results
from failures detected by the Java Virtual Machine, such as OutOfMemoryError and
normally should not be handled inside the application. The user-defined exceptions 
can be either checked or runtime. The Java specification~\cite{gosling2000java} 
suggests that user-defined exceptions should be checked (because doing so 
the callers of a method will know about the exceptions that a method can throw and so 
they can decide what to do about them). However, runtime exceptions can also be 
used to represent user-defined exceptional conditions.In fact, there is a long-lasting debate 
about the pros and cons of both approaches~\cite{javatut,stackoverlow,debate},
\footnote{152 questions in Stackoverlow are related to this debate}.

\begin{figure} \centering \includegraphics[width=\hsize]{new2_hierarchy.png}
  \caption{Exception Hierarchy in Java} \label{fig:exchier} \end{figure}

Besides, the runtime exceptions are also implicitly thrown by the runtime environment when a program violates 
the semantic constraints of the Java programming language (e.g., out-of-bounds array index, divide-by-zero 
error, null pointer references).\footnote{Some programming languages react to such errors by peremptorily terminating the program; 
other programming languages allow similar approaches e.g.C\#; and others let the program to continue
 its execution in some situations such as the out-of-bounds array index e.g., C++. }  
The programs are not expected to handle such runtime exceptions signaled by the runtime environment ~\cite{gosling2000java}.

@@

This exploration is guided by the set of best practices compiled in this study; we investigate whether characteristics 
of exception stack traces can pinpoint best practices violations. As mentioned before, 
call ``fragilities" characteristics on the exception-related code that favor the introduction
of failures such as uncaught exceptions and unintended handling.

@@

\noindent \fbox{ \begin{minipage}{0.96\columnwidth} \emph{
A multitude of programming mistakes were reported as the main causes of exception stack traces.
51,96\% of all projects reported at least one exception
 stack trace on which the NullPoiterException
was the root cause.
}
\end{minipage}}

@@

We can also observe that most of the reported exceptions were RuntimeExceptions
(64,85\%); and most of them were signaled on methods defined on the Application 
the Android Platform or on libraries.Taking into account the results of empirical study on 
the prevalende of ``programming erros" we know that a considerable ammount 
of such exceptions, are related to the programming mistakes
as pointed out previously.

However, within this set of runtime exceptions there are also the ones that do not correspond to programming mistakes,
and according to best practice IV should be documented as part of the exception interface of libraries/framework reusable
methods.

To investigate the obedience to such practice, we firstly filtered out all the exceptions implicitly
 signaled by  runtime environment (due to programming mistakes) since these exceptions 
should not be documented on the method signature.

Then we inspected the code each method (defined either on Android  Aplication Framework or on libraries) 
explicitly signaling a runtime exception not-related to programming mistakes.
As a result of this inspection, we could observe that only 1 method out of 118 inspected methods
 (i.e., 0,8\%) documented the explicitly thrown runtime exception in a Javadoc comment; and none of these methods
included the runtime exception  (reported on the stack) as part of the exception interface of the method (i.e., using 
throws clause on method signature). This result is aligned with the results of other study conducted by from 
Sacramento et al ~\cite{sacramento2006unchecked} which observed that the
runtime exceptions in .Net programs are most often not documented.

Such undocumented runtime exceptions represent a threat to system robustness, specially
when such exceptions are thrown then the third party code invoked inside the application 
(e.g. libraries, or framework utility code). In such cases the client usually do not have access to 
the source code. Hence in the absence of the exception documentation it is very difficult or even impossible
 for the client to design the application to deal with ``unforeseen runtime exceptions. As a consequence, the
 undocumented runtime exception may remain uncaught and lead to system crashes.

@@

\subsection{What can the Exception Wrappings tell us?}
%SELECT COUNT(*) FROM STACKTRACESUMMARY WHERE
%repo !='android'  AND ROOT_EXTENDS LIKE 'CHECKED%' AND EXCEPTION_EXTENDS LIKE 'RUNTIME'

%SELECT COUNT(*) FROM STACKTRACESUMMARY WHERE
%repo !='android'  AND ROOT_EXTENDS LIKE 'CHEC%' AND EXCEPTION_EXTENDS LIKE 'RUN%';

%SELECT COUNT(*) FROM STACKTRACESUMMARY WHERE
%repo !='android'  AND ROOT_EXTENDS LIKE 'CHEC%' AND EXCEPTION_EXTENDS LIKE 'RUN%'
%and SIGNALER_FROM LIKE 'LIB%';

 %Runtime &  Checked   & 70  / 18 & 114 /  34 &  66 /  9  &  0 /  & 19 / 19 &  29 / 6 \\
      %Runtime   &  Error   & 38  \ 8  & 57    &  50 \ 8  &  0   & 6 \ 2  &  1   /  0\\
%      Checked &  Runtime   & 12 \ 5  & 22 \ 9 & 4 &  0  & 9 \7 &  9 \ 2 \\
%      Checked & Error      & 7 \ 1 &  8 & 5  &  0  & 0 \ 1 &  3 \0 \\
 %     Error & Checked      & 4 \ 10  &  7 & 0  \ 6  &  7  & 0 \6 &  0   \ 8    \\
%      Error & Runtime     & 6 \ 2  &  13   & 1 &  1  & 0 \ 1 &  11 \ 3    \\

@@

  \hline
      %none   & Runtime   &  2360 & 359 \\
      %none  &   Checked  & 422  & 117 \\
      %none  &   Error   & 381  & 141 \\
      %none &  Undefined  &  15   & 10  \\
      %none & Throwable  &  1    & 1 \\
     %\hline
      %Runtime   & Runtime & 560  & 182 \\    
      %Checked   & Checked & 98   & 42  \\
      %Error     & Error   & 15   & 14  \\
    %\hline
  \end{tabular}
\caption{Wrappings comprising different exception types.}
\label{tab:wrappingandroid}
\end{table*}
Java is the only language that provides a hybrid exception model 
which offers three kinds of exceptions each one holding an inttended exception behavior (i.e., error, 
runtime and checked). The best practices I and II states the behavior that Error an Checked exceptions hold, which are: 
(i) Errors should not be handled inside the system since they usually represent unrecoverable conditions detected 
by the runtime environment (e.g.,  OutOfMemoryError); (ii) checked exceptions, on the other hand, should represent recoverable conditions.

Table~\ref{tab:wrappingandroid} presents all wrappings found in this study, that
envolve different exception types (i.e., Error, Checked, Exception). We call 
such a wrapping a ``cross-type wrapping".  Some of these wrappings may reveal 
violations to best practices I and II (e.g. Error wrapping a Checked exception) as discussed next.

Inspectioning the source code, of the methods involved in this kind of wrapping,
revealed an unexpected fragility: a checked exception thrown by a native method and not
declared on the exception interface of these methods signaling them. Such native method 
was defined on the Android Platform -  which uses Java Native Invocation (JNI) to access 
native C/C++ code.

The code snippet in Figure ~\ref{fig:snippets} (d) illustrates this fragility.
 The java/lang/Class declares a native method named getDeclaredMethods. 
The Java-side declaration of of this method does not have any throws clause, 
leading programmers and the compiler to think that no checked exceptions can be thrown.
 However, the C-code implementation did throw a "checked exception" NoSuchMethodException, 
violating the declaration. The Java compiler, cannot detect this violation because it does 
not perform static exception checking on native methods. This type of bug is hard to diagnose
because the developer usualy do not have access to the native implementations. 
Consequently, since it is not expected by the programmer, when such method throws 
this exception, such undocumented exception may remain
uncaught and cause the app to crash, or may be mistakenly handled by subsumption - 
The exception stack traces reporting this scenario is actually a real bug of Android 
Gingerbread version (which still accounts for 13.6\% of devices running Android).

Among the Android projects on which this exception stack trace was reported we can cite:
PageTurner (206 stars, 130 forks, 1184 commits), ActionBarSherlock (6,041 stars, 
3,436 forks, and 1479 commints), otto (1,618 stars,	342 forks, 	177 commints).

\noindent \fbox{ \begin{minipage}{0.96\columnwidth} \emph{
An unexpected fragility was detected: a checked exception being thown and not specified on 
on the ``exception interface" of the signaler methods.
}
\end{minipage}}

\emph{\textbf{Runtime Exception wrapping Error.}} We could find  64 exception stack traces 
on 46 projects on which an Error was wrapped in a runtime exception.
Manually inspecting each exception stack trace, we could observe that 44% of the wrappings were performed 
by android.os.AsyncTask. This class is responsible for dealing with asynchronous tasks in Android.
The code snippet is illustrated in Figure ~ref{wrappings} (c) shows that any instance of Throwable signaled in the context of an asynchronous 
task is converted in to a runtime and re-thrown.  Table ~\ref{tab:exampeswrap} exemplifies some of such wrappings.
Such wrapping masks the unrecoverable exception in a general excepion - doing so  that can be mistakenly handled

 %\begin{tabular}{|p{7.9cm} p{0.1cm}|}
\begin{tabular}{ll}
    \hline
 id & \bfseries{Runtime-Error Wrapping Examples}    \\  %\bfseries{\#}
    \hline
1 & java.lang.RuntimeException - java.lang.OutOfMemoryError  \\ %30
2 &java.lang.RuntimeException - java.lang.NoClassDefFoundError \\ %6
3 &java.lang.RuntimeException - java.lang.UnsatisfiedLinkError)  \\ %5
4& java.lang.RuntimeException -  java.lang.StackOverflowError)   \\ %4
\hline
 & \bfseries{Checked-Runtime Wrapping Examples}   \\
 \hline
5& ...android.robospice.CacheSavingException - java.lang.ClassCastException   \\ %6
6&java.lang.reflect.InvocationTargetException - java.lang.NullPointerException    \\ %3
7&java.lang.reflect.InvocationTargetException - ...ResourcesdNotFoundException  \\ %2
8&java.sql.SQLException - android.database.sqlite.SQLiteDiskIOException   \\ %2
 \hline
& \bfseries{Checked-Error Wrapping Examples}  \\
 \hline
9&java.lang.reflect.InvocationTargetException - java.lang.OutOfMemoryError  \\ %5
10&java.lang.Exception - java.lang.OutOfMemoryError   \\ %1
11&java.lang.reflect.InvocationTargetException - java.lang.NoSuchMethodError 	 \\ %1
12&java.lang.reflect.InvocationTargetException - java.lang.UnsatisfiedLinkError	 \\ %1
\hline
& \bfseries{Error-Checked Wrapping Examples}  \\
 \hline
13&java.lang.NoClassDefFoundError - java.lang.ClassNotFoundException   \\ %4
14&java.lang.AssertionError - javax.crypto.ShortBufferException)   \\ %3
\hline
& \bfseries{Error-Runtime Wrapping Examples}    \\
 \hline
15&java.lang.ExceptionInInitializerError - java.lang.NullPointerException   \\ %9
16&java.lang.ExceptionInInitializerError - java.lang.IllegalArgumentException 	 \\ %2
 \hline
  \end{tabular}
\caption{Examples of Cross-type wrappings}
\label{tab:exampeswrap}
\end{table}

\emph{\textbf{Checked Exception wrapping Error.}} Almost all of these wrappings
 were also caused by the reflection library used by Android applications. The methods responsible for the wrappings
were also native methods written in C. Table~\ref{tab:exampleswrap} illustrates
some oth these wrappings some of them are masking an OutOfMemoryError
into a checked exception.

\emph{\textbf{Error wrapping Checked Exception.}} These wrappings were performed
either by native methods defined by java.lang.Class (e.g., getDeclaredFields, getDeclaredConstructors) 
or methods from javax.crypto package. Table~\ref{tab:exampleswrap} illustrates
the exception types involved in such wrappings.

\emph{\textbf{Error wrapping Runtime.}} After manually inspecting the source code associated to such wrappings 
we discovered that all of them were caused by static initializers. If any exception is thrown in the context of a static initializer (i.e., static block) 
it is converted into an ExceptionInitializerError on the point where the class is first used. Table ~\ref{exampleswrap} illustrates
the exception types involved in such wrappings.

@@

Although some of these wrappings may be a result of design decisions,
the mis-use of exception wrapppings may make the exception handling 
code more complex (e.g., the multiple wrappings) and error-prone,
 and lead to what we call ``exception confusion problem"

To illustrate this problem we can use one of the wrappings discussed above.
When the developer is confronted with a checked exception, the designer of the API is telling him 
to handle the exceptional condition (according to Java Specification and 
best practices). However, if the developer tries to handle such exception,
which is wrapping (or ``hiding") an OutOfMemoryError, such handling 
may lead the program to an unpredictable state.

The cross-type wrappings show us that even if the developer sticks to Java
specification and best practices, it will be dificult for her/him to prevent the code from 
incurring into a situation as the one described above.

\noindent \fbox{ \begin{minipage}{0.96\columnwidth} \emph{
Some of the detected cross-type exception wrappings points to a potential ``exception confusion problem".
To illustrate: a checked exception wrapping an OutOfMemoryError; and the multiple wrapping Runtime-Checked-Runtime-Checked-Runtime-Checked-Runtime.
}
\end{minipage}}

@@

One may think that the solution for the uncaught exceptions may be to define a general handler, 
which is responsible for handling any exception that is not
adequately handled inside the applications. Although this 
solution may prevent  the system of abruptly crashing,
 such general handler will not have enough
contextual information to adequately handle the exception, 
and only tasks possible are: to present a message to the user
 and restart the application.

However, such handler cannot replace a carefully designed exception 
handling policy, which can hardly be designed in the absence of 
third-party documentation on the exceptions that
may the thrown by it.

@@

\subsection{On the null pointer problem}
The null references was firstly introduced by Tony Hoare in ALGOL W, which after some years he called his “one-billion-dollar mistake” ~\cite{hoare}:

\emph{``I call it my billion-dollar mistake. It was the invention of the null reference in 1965 [...] I couldn’t resist the temptation to put 
in a null reference, simply because it was so easy to implement. This has led to innumerable errors, vulnerabilities, and system 
crashes, which have probably caused a billion dollars of pain and damage in the last forty year."}

In this sudy, the null references were, infact, responsible for several reported application errors - providing more evidences to Hoare's speach.
This observation emphasizes the importance of solutions to avoid NullPointerExceptions, such as:
(i) lightweight intra-method null pointer analysis supported by Java8 @Nullable annotations ~\footnote{The @Nullable annotations are used by tools such as FindBugs, Eclipse, NetBeans, IntelliJ to detect potential 
null pointer dereferences at compile time. In April/2014 the Android Studio 0.5.5 was 
released with support to these annotations};
(ii) inter-method null pointer analysis tools such as the one proposed by Nanda and Sinha ~\cite{nanda2009accurate};
or (iii) whether language designs which avoid null pointers, such 
as Monads ~\cite{Walde95} (i.e., used in functional languages for values that may not be available 
or computations that may fail) could improve the robustness of Java programs.

@@

\noindent\emph{Internal Validity.} We used a heuristics-based parser to mine
exceptions from issues.  Our parsing strategy was conservative by default; for
example, we only considered exception names using a fully qualified class name
as valid exception identifiers, while, in many cases, developers use the
exception name in issue description. Conservative parsing may minimize false
positives, which was our initial target, but also tends to increase false
negatives, which means that some cases may have not been identified as
exceptions or stack traces. Our limited manual inspection did not reveal such
cases. Moreover, in this study we manually classified the exceptions 
related to programming mistakes. To ensure the quality of the analysis, 
we calculated the interrater agreement after three independent 
coders classified a randomly selected sample (of 25 exception 
types from the total 100); the interater agreement was high (96\%). 
Another threat relates to the fact that parts of our analysis 
are based on the availability of stack traces on issues on Github and Googlecode. 
In using these datasets, we make an underlying assumption: the stack traces reported on issues are 
representative of valid crash information of the applications. 
One way to mitigate such threat would be to access to the full 
set of crash data per application. Although some services exist 
to collect crash data from mobile applications ~\cite{BugSe14,BugSn14,Googl14,Acra14},
they do not provide open access to the crash reports of their client applications.
In our study, we mitigated this threat by manually inspecting
the source code associated to a subset of the reported exception stack traces.
Such subset comprise the stack traces related to the main findings 
of the study (e.g., `ùndocumented runtime and checked exceptions",
and cross-type wrappings").

>>>>

Along the recent years, we have witnessed an astonishing increase in the number of
mobile applications. If on the one hand such applications extend phones capabilities 
far beyond of the basic calls and textual messages, on the other hand
they have to face an increasing number of threats to application robustness
 arising from: faults on underlying middleware and hardware (e.g., camera, sensors);
faults on third party services and libraries; compatibility issues ~\cite{McDon13}; 
 memory and battery restrictions; noisy external resources ~\cite{Zhang12} (e.g., wireless connections,GPS, bluetooth) among others.

@@

Studies have shown, however, that the exception-related code is generally poorly understood and the
 less tested parts of the system~\cite{miller1997issues,Robil00,shah2010understanding, 
garcia2007extracting,garcia2001comparative,cabral2007exception,coelho2011unveiling}.
As a consequence they may affect the system the other way around.
The exception-related code may favor the introduction of failures such as 
uncaught exceptions~\cite{jo2004uncaught, Zhang12} - 
which can lead to system crashes, making the system even less robust~\cite{coelho2011unveiling}.

@@

       \item Aproximately 52\% of the reported stack traces were associated to programming mistackes. 
And 27,71\% of all stack traces of reported a java.lang.NullPointerException the root cause.

@@

Such outcomes point to threats not only to the developement of robust Android apps, 
but also to the developement of any robust Java-based system. For instance, in the absence of the 
``exception specification" of  a third-party code, it is difficult or 
even enfeasible for the developer to protect the code against ``unforeseen" exceptions. 
Since, in such cases the client usually do not have access to the source code, such undocumented exceptions may 
remain uncaught and lead to system crashes. Moreover, the cross-type exception wrappings 
(e.g., OutOfMemoryError is wrapped in a checked Exception) 
may lead to unintended handling problems - e.g, trying to handle an instance of OutOfMemoryError ``hiden" in 
a checked exception may lead the program to an unpredictable state.

@@

The Android Application Framework supports an event-driven development model. 
Among the basic elements that comprise an Android application, we can cite the
 Activity class. The Activity is the main class of an Android app (there is usualy one
Activity per application window). Each Activity defines a set of hook methods  
(e.g., OnCreate(), onPause()) that are dispatched automatically by the Android 
framework (by inversion of control), in response to infrastructure and application events.
The Android Application Framework also provides several supporting APIs, such as 
the APIs to support concurrent development (e.g. AsyncTask). 
Such APIs particularly useful due to the high demand for responsiveness 
in mobile environment (mobile apps can easily become 
unresponsive due to resource limitation and excessive network access).

@@

In Java, exceptions are represented according to a class hierarchy, on which
 every exception is an instance of the Throwable class, and can be of three kinds: the checked exceptions
(extends Exception), the runtime exceptions (extends RuntimeException) and errors
(extends Error)~\cite{gosling2000java}. The checked exception received such name
 because they must be declared on the method's exception interface (i.e., a list of exceptions that a method 
might raise during its execution) and the compiler statically checks if
 appropriate handlers are provided within the system.
Both runtime exceptions and errors are also known as unchecked exceptions, because 
they do not need to be specified on method exception interface and do not trigger any 
compile time checking.

By convention an Error represents an unrecoverable condition which usually results
from failures detected by the Java Virtual Machine, such as OutOfMemoryError and
normally should not be handled inside the application. The user-defined exceptions 
can be either checked or runtime. The Java specification~\cite{gosling2000java} 
suggests that user-defined exceptions should be checked (because doing so 
the callers of a method will know about the exceptions that a method can throw and so 
they can decide what to do about them). However, runtime exceptions can also be 
used to represent user-defined exceptional conditions.In fact, there is a long-lasting debate 
about the pros and cons of both approaches~\cite{javatut,stackoverlow,debate}
%\footnote{152 questions in Stackoverlow are related to this debate}.

The runtime exceptions are also implicitly thrown by the runtime environment when a program violates 
the semantic constraints of the Java programming language (e.g., out-of-bounds array index, divide-by-zero 
error, null pointer references).\footnote{Some programming languages react to such errors by peremptorily terminating the program; 
other programming languages allow similar approaches e.g.C\#; and others let the program to continue
 its execution in some situations such as the out-of-bounds array index e.g., C++. }  
The programs are not expected to handle such runtime exceptions signaled by the runtime environment ~\cite{gosling2000java}.

%\begin{figure} \centering \includegraphics[width=\hsize]{new2_hierarchy.png}
% \caption{Exception Hierarchy in Java} \label{fig:exchier} \end{figure}

@@

This exploration is guided by the set of best practices compiled in this study; we investigate whether characteristics 
of exception stack traces and the related source code it can pinpoint best practices violations. As mentioned before, 
we call ``fragilities" characteristics on the exception-related code that favor the introduction
of failures such as uncaught exceptions and unintended handling.

@@

\noindent \fbox{ \begin{minipage}{0.96\columnwidth} \emph{
Aproximately 52\% of the reported stack traces were associated to programming mistackes. 
And 27,71\% of all stack traces of reported a java.lang.NullPointerException the root cause.
}
\end{minipage}}

@@

We can also observe that most of the reported exceptions were RuntimeExceptions
(64,85\%); and most of them were signaled on methods defined on the Application 
the Android Platform or on libraries.Taking into account the results of empirical study on 
the prevalende of ``programming erros" we know that a considerable ammount 
of such exceptions, are related to the programming mistakes
as pointed out previously. However, within this set of runtime exceptions there are also the ones that do not correspond to programming mistakes,
and according to Java specification should be documented as part of the exception interface of libraries/framework reusable
methods.

\emph{\textbf{Inspecting Exception Interfaces.}} To investigate the obedience to such practice, we firstly filtered out all the exceptions implicitly
 signaled by  runtime environment (due to programming mistakes) since these exceptions 
should not be documented on the method signature.  Then we inspected the code each method 
(defined either on Android  Aplication Framework or on libraries) 
explicitly signaling a runtime exception not-related to programming mistakes.
As a result of this inspection, we could observe that only 1 method out of 118 inspected methods
 (i.e., 0,8\%) documented the explicitly thrown runtime exception in a Javadoc comment; and none of these methods
included the runtime exception  (reported on the stack) as part of the exception interface of the method (i.e., using 
throws clause on method signature). This result is aligned with the results of other study conducted by from 
Sacramento et al ~\cite{sacramento2006unchecked} which observed that the
runtime exceptions in .Net programs are most often not documented.
These ``undocumented runtime exceptions" represent a threat to system robustness, specially
when such exceptions are thrown then the third party code invoked inside the application 
(e.g. libraries, or framework utility code). In such cases the client usually do not have access to 
the source code. Hence in the absence of the exception documentation it is very difficult or even impossible
 for the client to design the application to deal with ``unforeseen runtime exceptions. As a consequence, the
 undocumented runtime exception may remain uncaught and lead to system crashes.

@@

\emph{\textbf{On the Presence of Checked Exceptions on Exception Interfaces.}} 
In this study, the exception stack trace analysis revealed an unexpected fragility: a 
checked exception thrown by a native method and not declared on the exception 
interface of these methods signaling them. Such native method was defined on
 the Android Platform -  which uses Java Native Invocation (JNI) to access 
native C/C++ code.

This exceptions was thrown by the method getDeclaredMethods defined in java.lang.Class. 
The Java-side declaration of this method does not have any throws clause, 
leading programmers and the compiler to think that no checked exceptions can be thrown.
 However, the C-code implementation did throw a "checked exception" NoSuchMethodException, 
violating the declaration. The Java compiler, could not detect this violation, because it does 
not perform static exception checking on native methods. This type of bug is hard to diagnose
because the developer usualy do not have access to the native implementations. 
Consequently, since it is not expected by the programmer, when such method throws 
this exception, such undocumented exception may remain
uncaught and cause the app to crash, or may be mistakenly handled by subsumption - 
The exception stack traces reporting this scenario is actually a real bug of Android 
Gingerbread version (which still accounts for 13.6\% of devices running Android).
%Among the Android projects on which this exception stack trace was reported we can cite:
%PageTurner (206 stars), ActionBarSherlock (6,041 stars), otto (1,618 stars).

\noindent \fbox{ \begin{minipage}{0.96\columnwidth} \emph{
An unexpected fragility was detected: a checked exception being thown and not specified  
on the ``exception interface" of the signaler methods.
}
\end{minipage}}

\subsection{What can the Exception Wrappings tell us?}

 %\begin{tabular}{|p{7.9cm} p{0.1cm}|}
\begin{tabular}{ll}
    \hline
 id & \bfseries{Runtime wrapping Error}    \\  %\bfseries{\#}
    \hline
1 & java.lang.RuntimeException - java.lang.OutOfMemoryError  \\ %30
2& java.lang.RuntimeException -  java.lang.StackOverflowError)   \\ %4
\hline
& \bfseries{Checked wrapping Error}  \\
 \hline
3&java.lang.reflect.InvocationTargetException - java.lang.OutOfMemoryError  \\ %5
4&java.lang.Exception - java.lang.OutOfMemoryError   \\ %1
\hline
& \bfseries{Error wrapping Checked}  \\
 \hline
5&java.lang.NoClassDefFoundError - java.lang.ClassNotFoundException   \\ %4
6&java.lang.AssertionError - javax.crypto.ShortBufferException)   \\ %3
\hline
& \bfseries{Error wrapping Runtime}    \\
 \hline
7&java.lang.ExceptionInInitializerError - java.lang.NullPointerException   \\ %9
8&java.lang.ExceptionInInitializerError - java.lang.IllegalArgumentException 	 \\ %2
 \hline
  \end{tabular}
\caption{Examples of Cross-type wrappings}
\label{tab:exampeswrap}
\end{table}

@@

  \hline
      %none   & Runtime   &  2360 & 359 \\
      %none  &   Checked  & 422  & 117 \\
      %none  &   Error   & 381  & 141 \\
      %none &  Undefined  &  15   & 10  \\
      %none & Throwable  &  1    & 1 \\
     %\hline
      %Runtime   & Runtime & 560  & 182 \\    
      %Checked   & Checked & 98   & 42  \\
      %Error     & Error   & 15   & 14  \\
    %\hline
  \end{tabular}
\caption{Wrappings comprising different exception types.}
\label{tab:wrappingandroid}
\end{table*}

Java is the only language that provides a hybrid exception model 
which offers three kinds of exceptions each one holding an intended exception behavior (i.e., error, 
runtime and checked). The best practices I and II are related to such type-related-behavior: 
(i) Errors should not be handled inside the system, since they usually represent unrecoverable conditions detected 
by the runtime environment (e.g.,  OutOfMemoryError); (ii) checked exceptions, on the other hand, should represent 
recoverable conditions.

Table~\ref{tab:wrappingandroid} presents the wrappings found in this study, that
envolve different exception types (i.e., Error, Checked, Exception). We call 
such a wrapping a ``cross-type wrapping". Next we exemplify some of such
 wrappings and discuss why some of them may negatively impact the robustness.

\emph{\textbf{Runtime Exception wrapping Error.}} We could find  64 exception stack traces 
on 46 projects on which an Error was wrapped in a runtime exception.
Manually inspecting each exception stack trace, we could observe that 44\% of the wrappings were performed 
by android.os.AsyncTask. This class is responsible for dealing with asynchronous tasks in Android.
The code snippet is illustrated in Figure ~ref{wrappings} (c) shows that any instance of Throwable signaled 
in the context of an asynchronous task is converted in to a runtime and re-thrown.  Table ~\ref{tab:exampeswrap} exemplifies some of such wrappings.
Such wrapping masks an unrecoverable error in a general exception - doing so it that can be mistakenly handled.

\emph{\textbf{Checked Exception wrapping Error.}} Most of these wrappings
 were also caused by the reflection library used by applications' methods. The methods responsible for the wrappings
were also native methods written in C. Table~\ref{tab:exampleswrap} illustrates
some oth these wrappings some of them are masking an OutOfMemoryError
into a checked exception.

\emph{\textbf{Error wrapping Runtime and Checked Exceptions.}} 
These wrappings were performed by: native methods defined by 
java.lang.Class (e.g., getDeclaredFields, getDeclaredConstructors), 
methods from javax.crypto package, and Java static initializers. 
If any exception is thrown in the context of a static initializer 
(i.e., static block)  it is converted into an ExceptionInitializerError 
on the point where the class is first used.  Table~\ref{tab:exampleswrap} illustrates
examples of such wrappings.

@@

Although some of these wrappings may be a result of design decisions,
the mis-use of exception wrapppings may make the exception handling 
code more complex (e.g., the multiple wrappings) and error-prone,
 and lead to what we call ``exception confusion problem". To illustrate this problem we can use one of the wrappings discussed above.
When the developer is confronted with a checked exception, the designer of the API is telling him 
to handle the exceptional condition (according to Java Specification and 
best practices). However, if the developer tries to handle such exception,
which is wrapping (or ``hiding") an OutOfMemoryError, such handling 
may lead the program to an unpredictable state.

\noindent \fbox{ \begin{minipage}{0.96\columnwidth} \emph{
The cross-type wrappings show us that even if the developer sticks to Java
specification and best practices, it will be dificult for her/him to prevent the code from 
incurring into situations that lead the program to an unpredictable state. 
}
\end{minipage}}

@@

One may think that the solution for the uncaught exceptions may be to define a general handler, 
which is responsible for handling any exception that is not
adequately handled inside the applications. Although this 
solution may prevent  the system of abruptly crashing,
 such general handler will not have enough
contextual information to adequately handle the exception, 
and only tasks possible are: to present a message to the user
 and restart the application. However, such handler cannot replace a carefully designed exception 
handling policy ~\cite{Robil00}, which can hardly be designed in the absence of 
third-party documentation on the exceptions that
may the thrown by it.

@@

\subsection{On the null pointer problem}
The null references was firstly introduced by Tony Hoare in ALGOL W, which after some years he called his ``one-billion-dollar mistake"~\cite{hoare2}:
\emph{``I call it my billion-dollar mistake. It was the invention of the null reference in 1965 [...] This has led to innumerable errors, vulnerabilities, and system 
crashes, which have probably caused a billion dollars of pain and damage in the last forty year."}

In this sudy, the null references were, infact, responsible for several reported application errors - providing more evidences to Hoare's speach.
This observation emphasizes the importance of solutions to avoid NullPointerExceptions, such as:
(i) lightweight intra-method null pointer analysis supported by Java8 @Nullable annotations~\footnote{Such annotations are used by tools e.g., Eclipse, IntelliJ to detect potential 
null pointer dereferences at compile time. The Android Studio 0.5.5 released in April 2014 support to these annotations.};
(ii) inter-method null pointer analysis tools such as the one proposed by Nanda and Sinha ~\cite{nanda2009accurate};
or (iii) whether language designs which avoid null pointers, such 
as Monads ~\cite{Walde95} (i.e., used in functional languages for values that may not be available 
or computations that may fail) could improve the robustness of Java programs.

@@

\noindent\emph{Internal Validity.} We used a heuristics-based parser to mine
exceptions from issues.  Our parsing strategy was conservative by default; for
example, we only considered exception names using a fully qualified class name
as valid exception identifiers, while, in many cases, developers use the
exception name in issue description. Conservative parsing may minimize false
positives, which was our initial target, but also tends to increase false
negatives, which means that some cases may have not been identified as
exceptions or stack traces. Our limited manual inspection did not reveal such
cases. Moreover, in this study we manually classified the exceptions 
related to programming mistakes. To ensure the quality of the analysis, 
we calculated the interrater agreement after three independent 
coders classified a randomly selected sample (of 25 exception 
types from the total 100); the interater agreement was high (96\%). 
Another threat relates to the fact that parts of our analysis 
are based on the availability of stack traces on issues on Github and Googlecode. 
In using these datasets, we make an underlying assumption: the stack traces reported on issues are 
representative of valid crash information of the applications. 
One way to mitigate such threat would be to access to the full 
set of crash data per application. Although some services exist 
to collect crash data from mobile applications ~\cite{BugSe14,BugSn14,Googl14,Acra14},
they do not provide open access to the crash reports of their client applications.
In our study, we mitigated this threat by manually inspecting
the source code associated to a subset of the reported exception stack traces.
Such subset comprise the stack traces related to the main findings 
of the study (e.g., ``undocumented runtime and checked exceptions",
and cross-type wrappings").
